---
title: "Introduction to Reshape"
author: 
    name: "Monique Hernandez"
    affiliation: University of California, Santa Barbara
date: "Augustt 18, 2026"
format:
  html:
    toc: true
execute:
  message: false
  warning: false
---
## Day 7 PM Live Coding

```{r, message=FALSE}
library(tidyverse)

surveys <- read_csv("data/surveys.csv") # read data files for surveys
species <- read_csv("data/species.csv") # read data files for species
plots <- read_csv("data/plots.csv") # read data files for plots

```


# pivot_wider() ----------------------------------------------------------

## Finding weight by sex of species
```{r}
weight_by_sex <- surveys |> # new data weight_by_sex
  filter(!is.na(sex)) |> # filter by sex removing NA
  summarize( # summarize function, return a new row
    mean_weight = mean(weight, na.rm = TRUE), # name mean_weight 
    .by = c(species_id, sex) # by sex of species id
  )
```

## Using the pivot wider for new data set

```{r}
weight_by_sex_wider <- weight_by_sex |> # new data weight_by_sex_wider
  pivot_wider( # pivot_wider function
    names_from = sex, # from sex
    values_from = mean_weight # values from the mean_weight
  )
weight_by_sex_wider # call new data
```

::: {.callout-note}
## pivot_wider
pivot_wider is useful when one column contains multiple sets of values that you want to use with each other
(ex. calculating weight difference between M and F)
:::

# Another example
```{r}
taxa_by_year <- surveys |> # new data for taxa by years
  filter(year >= 1999) |> # filter by year
  # This join will add the taxa column from species to my surveys subset
  inner_join( # join function
    select(species, species_id, taxa), # select columns
    join_by(species_id) # join tables
  ) |>
  count(year, taxa) # count function 
taxa_by_year
```


# Pivot this wider, so we have columns for Birds and Rodents
```{r}
taxa_by_year_wider <- taxa_by_year |>
  pivot_wider(
    names_from = taxa,
    values_from = n
  )
taxa_by_year_wider
```

# pivot_longer() ---------------------------------------------------------

::: {.callout-note}
## Inverses
pivot_wider() and pivot_longer() are inverses of each other
:::
```{r}
taxa_by_year_wider
```

```{r}
taxa_by_year_wider|>
    pivot_longer( # pivot back to long format
    cols = Rodent:Bird,
    # where are the names going?
    names_to = "taxa",
    # where are the values going?
    values_to = "n"
  )

```